---
title: "100-alignment"
output: html_document
date: "2026-01-23"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

```{r}
library(tidyverse)
```

This file is for making notes about alignment and checking alignment success. I've put a basic alignment script that you can execute in a directory to align files.

```{r}
meta<-read_csv("metadata/meta01222025.csv")
meta %>% group_by(Species, `Tributary/Collection Location`) %>% summarize(Count=n())
```

Our sequences are labeled like this: plate-9_RA_GGGTGTTCTATGCAGG.fastq plate-9_RB_GGGTGTTCTATGCAGG.fastq

So to recreate the individual sequence names I'm going to do this.

```{r}
meta %>% mutate %>% mutate(Forward=paste0("plate-",PlateNumber,"_RA_GG",`Sample/Well Barcode (RAD Adaptor Seq)`,"TGCAGG.fastq")) %>%
  mutate(Reverse=paste0("plate-",PlateNumber,"_RB_GG",`Sample/Well Barcode (RAD Adaptor Seq)`,"TGCAGG.fastq")) %>%
  select(Forward, Reverse, Ind) %>% write_tsv("metadata/sequence-names.tsv") 
```

To set up alignment on FARM, we can symlink data from my home directory like so:

```{sh, eval=FALSE}
git clone https://github.com/jj-sweeney/newark 
cd newark
mkdir data
cd data
cut -f 1 ../metadata/sequence-names.tsv  | while read line; do ln -s /home/maccamp/fish-lake/data/align/$line ./ ; done;
cut -f 2 ../metadata/sequence-names.tsv  | while read line; do ln -s /home/maccamp/fish-lake/data/align/$line ./ ; done;
```

Now there are 496 forward (RA) and 496 reverse (RB) files in ./data/

Example execution in home/maccamp/newark/data

Changes in partitions, checking with /opt/hpccf/bin/slurm-show-resources.py --full

```{sh, eval=FALSE}
head -n 11 ../metadata/sequence-names.tsv | tail -n 10 > align-10.tsv
bash ../100.1-do-align.sh align-10.tsv /home/jsweeney/newark/genomes/GCA_042767315.1_fSipBic1.0.p_genomic.fna.gz
```

This is working. Joe - you can create a genomes/siphateles directory and build indices there and align!

01252026 adding some more notes.

On farm I have the job checking set up wiht an alias, 'sq' in my .bash_profile 'alias sq="squeue -u \$(whoami)"\`.

in newark/data on farm:

```{sh, eval=FALSE}
cat *.stats > align-10.csv
```

moving locally to outputs/100

```{r}
align<-read_csv("/Users/joesweeney/Desktop/outputs/100/align-10.csv", col_names = c("Ind","Reads2","Dedup2","Coverage2"))
# to compare against our previous alignments, I imagine this will be more instructive with the Siphateles genome
# Data frame could also be made tidy with "reference genome option"

align10<-meta %>% left_join(align) %>% arrange(-Dedup2) %>% filter(Dedup2 > 0)

# A tidy version to compare to two different references Here they will be the same but just an idea of something to try.
meta1<-meta %>% mutate(ReferenceGenome="Gila")
meta2<-meta %>% select(-Reads, -Dedup, -Coverage) %>% left_join(read_csv("/Users/joesweeney/Desktop/outputs/100/align-10.csv", col_names = c("Ind","Reads","Dedup","Coverage"))) %>% mutate(ReferenceGenome="Siphateles")
meta3<-bind_rows(meta1,meta2)
```

```{r}
ggplot(align10) +
  geom_histogram(aes(x=Dedup2, fill=`Tributary/Collection Location`)) +
  theme_bw() +
  theme(panel.grid=element_blank()) +
  scale_fill_viridis_d(option="H") +
  ylab("Count\n") +
  xlab("\nAligned Reads After Filtering")
  
```

```{r}
ggplot(meta3) +
  geom_histogram(aes(x=Dedup, fill=`Tributary/Collection Location`)) +
  theme_bw() +
  theme(panel.grid=element_blank()) +
  scale_fill_viridis_d(option="H") +
  ylab("Count\n") +
  xlab("\nAligned Reads After Filtering") +
  facet_wrap(.~ReferenceGenome)

```
