---
title: "100-alignment"
output: html_document
date: "2026-01-23"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

```{r}
library(tidyverse)
```

This file is for making notes about alignment and checking alignment success. I've put a basic alignment script that you can execute in a directory to align files.

```{r}
meta<-read_csv("metadata/meta01222025.csv")
meta %>% group_by(Species, `Tributary/Collection Location`) %>% summarize(Count=n())
```

Our sequences are labeled like this:
plate-9_RA_GGGTGTTCTATGCAGG.fastq
plate-9_RB_GGGTGTTCTATGCAGG.fastq

So to recreate the individual sequence names I'm going to do this.

```{r}
meta %>% mutate %>% mutate(Forward=paste0("plate-",PlateNumber,"_RA_GG",`Sample/Well Barcode (RAD Adaptor Seq)`,"TGCAGG.fastq")) %>%
  mutate(Reverse=paste0("plate-",PlateNumber,"_RB_GG",`Sample/Well Barcode (RAD Adaptor Seq)`,"TGCAGG.fastq")) %>%
  select(Forward, Reverse, Ind) %>% write_tsv("metadata/sequence-names.tsv") 
```

To set up alignment on FARM, we can symlink data from my home directory like so:

```{sh, eval=FALSE}
git clone https://github.com/jj-sweeney/newark
cd newark
mkdir data
cd data
cut -f 1 ../metadata/sequence-names.tsv  | while read line; do ln -s /home/maccamp/fish-lake/data/align/$line ./ ; done;
cut -f 2 ../metadata/sequence-names.tsv  | while read line; do ln -s /home/maccamp/fish-lake/data/align/$line ./ ; done;
```


Now there are 496 forward (RA) and 496 reverse (RB) files in ./data/ 

Example execution in home/maccamp/newark/data

Changes in partitions, checking with /opt/hpccf/bin/slurm-show-resources.py --full 

```{sh, eval=FALSE}
head -n 11 ../metadata/sequence-names.tsv | tail -n 10 > align-10.tsv
bash ../100.1-do-align.sh align-10.tsv /home/maccamp/genomes/gila-orcutti/GCA_026230005.1_fGilOrc1.0.hap1_genomic.fna.gz
```

This is working.  Joe - you can create a genomes/siphateles directory and build indices there and align!
