---
title: "101-angsd-basics"
output: html_document
date: "2026-02-16"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

```{r}
library(tidyverse)
library(snpR)
```

Setting up some examples to work with existing alignments.

I have 10 alignments in data from Cold Creak Reservoir in data. I'll set up some basics.  First getting the metadata.    

```{r}
#align10<-read_csv("../outputs/100/align-10.csv", col_names = c("Ind","Reads","Dedup","Cov"))
#meta<-read_csv("metadata/meta01222025.csv")
#m10<-left_join(align10,meta) %>% mutate(Path=paste0("data/",Ind,".sort.flt.bam")) %>% mutate(Pop=`Tributary/Collection Location`)
```

now using all 483 individuals (filtering for those with greater than 1E6 reads) -js20260225

```{r}
align496<-read_csv("../outputs/100/align-496.csv", col_names = c("Ind","Reads","Dedup","Cov"))
filtered_align <- subset(align496, Dedup > 100000)
meta<-read_csv("metadata/meta01222025.csv")
m483<-left_join(filtered_align,meta, by="Ind") %>%   mutate(Path=paste0("data/",Ind,".sort.flt.bam")) %>% mutate(Pop=`Tributary/Collection Location`)
```

Set up sample names and bamlist.

```{sh, eval=FALSE}
mkdir bamlists
```

```{r}
#m10 %>% select(Path) %>% write_tsv("bamlists/10.bamlist", col_names = FALSE)
#m10 %>% select(GVL_Code) %>% write_tsv("bamlists/10.names", col_names = FALSE)
```

```{r}
m483 %>% select(Path) %>% write_tsv("bamlists/483.bamlist", col_names = FALSE)
m483 %>% select(GVL_Code) %>% write_tsv("bamlists/483.names", col_names = FALSE)
```

The reference genome for these was `/home/maccamp/genomes/gila-orcutti/GCA_026230005.1_fGilOrc1.0.hap1_genomic.fna.gz` a large contig is `JAPDVR010000001.1`.       

The largest contig for the S. bicolor genome is `JBECYN010000001.1` - js20260224

On farm, can try something like this.

```{sh, eval=FALSE}
git fetch
git branch -v -a
git switch angsd-basics

mkdir outputs
mkidr outputs/101
```


Now I can try some basic angsd-y things. I'll use the module to see if it works.   

```{sh, eval=FALSE}
#module load angsd
#srun -t 2:00:00 -p bmh --account=millermrgrp --mem=128G --nodes=1 --ntasks-per-node=1 --cpus-per-task=12 angsd -nthreads 24 \
#-minInd 9 -bam bamlists/10.bamlist \
#-ref /home/jsweeney/newark/genomes/genome.fasta \
#-r JBECYN010000001.1 -out outputs/101/snps-10 \
#-minMaf 0.05 -minMapQ 20 -minQ 20 -GL 1 -doMajorMinor 1 -doMaf 1 -SNP_pval 1e-6 \
#-doGeno 2 -doPost 1 -postCutoff 0.95 -doPlink 2 > outputs/101/snps-10.out 2> outputs/101/snps-10.err &

```

	-> Number of sites retained after filtering: 2910 

```{sh, eval=FALSE}
module load angsd
srun -t 2:00:00 -p bmh --account=millermrgrp --mem=128G --nodes=1 --ntasks-per-node=1 --cpus-per-task=12 angsd -nthreads 24 \
-minInd 9 -bam bamlists/483.bamlist \
-ref /home/jsweeney/newark/genomes/genome.fasta \
-r JBECYN010000001.1 -out outputs/101/snps-483 \
-minMaf 0.05 -minMapQ 20 -minQ 20 -GL 1 -doMajorMinor 1 -doMaf 1 -SNP_pval 1e-6 \
-doGeno 2 -doPost 1 -postCutoff 0.95 -doPlink 2 > outputs/101/snps-483.out 2> outputs/101/snps-483.err &

```

-> Number of sites retained after filtering: 15062 

```{sh, eval=FALSE}
#convert to .vcf
module load plink
plink --tped outputs/101/snps-483.tped --tfam outputs/101/snps-483.tfam --out outputs/101/plink-binary --recode --allow-extra-chr --noweb
plink --ped outputs/101/plink-binary.ped --map outputs/101/plink-binary.map --recode vcf --allow-extra-chr -out outputs/101/plink
module load tabix
bgzip outputs/101/plink.vcf
tabix outputs/101/plink.vcf.gz
module load bcftools
bcftools reheader --samples bamlists/483.names outputs/101/plink.vcf.gz > outputs/101/renamed.vcf.gz
module load tabix
tabix outputs/101/renamed.vcf.gz
```

Now there is a .vcf file with the SNPs and names by sample names, putting in my outputs/101 locally.    

# snpR

Using "Pop" for convenience, but is sampling location.

```{r}
snps<-import.snpR.data(genotypes = "../outputs/101/renamed.vcf.gz", 
      sample.meta = m483 %>% select(Pop))
```

```{r}
pca<-plot_clusters(snps)
pca$plots
```


Can pull data frame for plotting from pca object.    

```{r}
df<-pca$data$pca
ggplot(df) +
  geom_point(aes(x=PC1, y=PC2, fill=Pop), pch=21) +
  theme_bw() +
  theme(panel.grid = element_blank()) +
  xlab(paste0("PC1 (",pca$pca_loadings[1],"%)")) +
  ylab(paste0("PC2 (",pca$pca_loadings[2],"%)")) +
  ggtitle("Super informative plot title\n") +
  theme(plot.title = element_text(hjust=0.5))

```

on farm, switching back to main branch
```{sh, eval=FALSE}
git fetch
git branch -v -a
git switch main
```