---
title: "101-angsd-basics"
output: html_document
date: "2026-02-16"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

```{r}
library(tidyverse)
```

Setting up some examples to work with existing alignments.

I have 10 alignments in data from Cold Creak Reservoir in data. I'll set up some basics.  First getting the metadata.    

```{r}
align10<-read_csv("outputs/100/align-10.csv", col_names = c("Ind","Reads","Dedup","Cov"))
meta<-read_csv("metadata/meta01222025.csv")
m10<-left_join(align10,meta) %>% mutate(Path=paste0("data/",Ind,".sort.flt.bam"))
```

Set up sample names and bamlist.

```{sh, eval=FALSE}
mkdir bamlists
```

```{r}
m10 %>% select(Path) %>% write_tsv("bamlists/10.bamlist", col_names = FALSE)
m10 %>% select(GVL_Code) %>% write_tsv("bamlists/10.names", col_names = FALSE)
```

The reference genome for these was `/home/maccamp/genomes/gila-orcutti/GCA_026230005.1_fGilOrc1.0.hap1_genomic.fna.gz` a large contig is `JAPDVR010000001.1`.       

On farm, can try something like this.

```{sh, eval=FALSE}
git fetch
git branch -v -a
git switch angsd-basics

mkdir outputs
mkidr outputs/101
```


Now I can try some basic angsd-y things. I'll use the module to see if it works.   

```{sh, eval=FALSE}
srun -t 2:00:00 -p bmh --mem=128G --nodes=1 --ntasks-per-node=1 --cpus-per-task=12 angsd -nthreads 24 \
-minInd 9 -bam bamlists/10.bamlist \
-ref /home/maccamp/genomes/gila-orcutti/GCA_026230005.1_fGilOrc1.0.hap1_genomic.fna.gz \
-r JAPDVR010000001.1 -out outputs/101/snps-10 \
-minMaf 0.05 -minMapQ 20 -minQ 20 -GL 1 -doMajorMinor 1 -doMaf 1 -SNP_pval 1e-6 \
-doGeno 2 -doPost 1 -postCutoff 0.95 -doPlink 2 > outputs/101/snps-10.out 2> outputs/101/snps-10.err &

```

