---
title: "101.2-genome-snp-discov"
author: "joe sweeney"
date: "2026-04-17"
output: html_document
---

```{r setup, include=FALSE} 
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(snpR)
```

taken from 101-angsd-basics
used "fil" instead of 404 so that you dont have to keep updating n with
  filtering adjustments
```{r}
align496<-read_csv("outputs/100/align-496.csv", col_names = c("Ind","Reads","Dedup","Cov"))
filtered_align <- subset(align496, Dedup > 500000)
filtered_align <- subset(filtered_align, Cov > 0.1)
meta<-read_csv("metadata/meta01222025.csv")
m_fil <-left_join(filtered_align,meta, by="Ind") %>%   mutate(Path=paste0("data/",Ind,".sort.flt.bam")) %>% mutate(Pop=`Tributary/Collection Location`)
m_nwrk <- subset(m_fil, Species == "newarkensis" )
```

make bamlists from filtered samples (fil) and just newark samples (nwrk)
```{r}
m_fil %>% select(Path) %>% write_tsv("bamlists/fil.bamlist", col_names = FALSE)
m_fil %>% select(GVL_Code) %>% write_tsv("bamlists/fil.names", col_names = FALSE)
m_nwrk %>% select(Path) %>% write_tsv("bamlists/nwrk.bamlist", col_names = FALSE)
m_nwrk %>% select(GVL_Code) %>% write_tsv("bamlists/nwrk.names", col_names = FALSE)
```

now doing snp discovery in RAD-projects/newark using all contigs 
  from s bicolor reference
```{sh, eval=FALSE}
mkdir outputs/101.2
n_ind=$(wc -l < bamlists/fil.bamlist)
min_ind=$(( n_ind * 9 / 10 ))
echo $min_ind
sbatch run_angsd.sh 
```

converting to vcf
```{sh, eval=FALSE}
module load plink
plink --tped outputs/101.2/snps-fil.tped --tfam outputs/101.2/snps-fil.tfam --out outputs/101.2/plink-binary --recode --allow-extra-chr --noweb
plink --ped outputs/101.2/plink-binary.ped --map outputs/101.2/plink-binary.map --recode vcf --allow-extra-chr -out outputs/101.2/plink
module load tabix
bgzip outputs/101.2/plink.vcf
tabix outputs/101.2/plink.vcf.gz
module load bcftools
bcftools reheader --samples bamlists/fil.names outputs/101.2/plink.vcf.gz > outputs/101.2/renamed_fil.vcf.gz
module load tabix
tabix outputs/101.2/renamed_fil.vcf.gz
```

filtering snps in LD
```{sh, eval=FALSE}
module load bcftools
bcftools +fill-tags outputs/101.2/renamed_fil.vcf.gz  -- -t MAF,F_MISSING | bcftools view -i 'F_MISSING < 0.05' | bcftools view -i 'INFO/MAF > 0.05' |  bcftools +prune -m 0.20 -w 50kb --nsites-per-win 1 --nsites-per-win-mode maxAF  > outputs/101.2/pruned.vcf

bcftools +fill-tags outputs/101.2/renamed_fil.vcf.gz  -- -t MAF,F_MISSING | bcftools view -i 'F_MISSING < 0.05' | bcftools view -i 'INFO/MAF > 0.05' |  bcftools +prune -m 0.20 -w 50kb --nsites-per-win 1 --nsites-per-win-mode rand  > outputs/101.2/pruned-rand.vcf
```

moving to local storage
```{sh, eval=FALSE}
gzip -k pruned.vcf
scp jsweeney@farm.hpc.ucdavis.edu:/group/ajfingergrp/RAD-projects/newark/outputs/101.2/pruned.vcf.gz
```

```{r}
snps<-import.snpR.data(genotypes = "outputs/101.2/pruned.vcf.gz", 
      sample.meta = m_fil %>% select(Pop))

pca<-plot_clusters(snps)
pca$plots
```

Creating a PCA for all inds by sample site
```{r}
df<-pca$data$pca
ggplot(df) +
  geom_point(aes(x=PC1, y=PC2, fill=Pop), pch=21) +
  theme_bw() +
  theme(panel.grid = element_blank()) +
  xlab(paste0("PC1 (",pca$pca_loadings[1],"%)")) +
  ylab(paste0("PC2 (",pca$pca_loadings[2],"%)")) +
  ggtitle("Super informative plot title\n") +
  theme(plot.title = element_text(hjust=0.5))
```

now all inds by species
```{r}
snps_by_sp<-import.snpR.data(genotypes = "outputs/101.2/pruned.vcf.gz", 
      sample.meta = m_fil %>% select(Species))
pca_sp<-plot_clusters(snps_by_sp)
pca_sp$plots
```

```{r}
df_sp<-pca_sp$data$pca
ggplot(df_sp) +
  geom_point(aes(x=PC1, y=PC2, fill=Species), pch=21) +
  theme_bw() +
  theme(panel.grid = element_blank()) +
  xlab(paste0("PC1 (",pca$pca_loadings[1],"%)")) +
  ylab(paste0("PC2 (",pca$pca_loadings[2],"%)")) +
  ggtitle("SNP clustering by species\n") +
  theme(plot.title = element_text(hjust=0.5))
```

```{r}
snps_n<-import.snpR.data(genotypes = "outputs/101.2/pruned.vcf.gz", 
      sample.meta = m_fil %>% select(Species, Pop))

pca_n<-plot_clusters(snps_n)
pca_n$plots
```

```{r}
df_n<-pca_n$data$pca %>% filter(Species == "newarkensis")
ggplot(df_n) +
  geom_point(aes(x=PC1, y=PC2, fill=Pop), pch=21) +
  theme_bw() +
  theme(panel.grid = element_blank()) +
  xlab(paste0("PC1 (",pca$pca_loadings[1],"%)")) +
  ylab(paste0("PC2 (",pca$pca_loadings[2],"%)")) +
  ggtitle("SNP clustering by pop (newarkensis)\n") +
  theme(plot.title = element_text(hjust=0.5))

```

--------------------------------------------------------------------------------
now doing snp discovery on only newarkensis
```{sh, eval=FALSE}
n_ind=$(wc -l < bamlists/nwrk.bamlist)
min_ind=$(( n_ind * 9 / 10 ))
echo $min_ind
sbatch run_angsd.sh 
```

```{sh, eval=FALSE}
module load plink
plink --tped outputs/101.2/snps-nwrk.tped --tfam outputs/101.2/snps-nwrk.tfam --out outputs/101.2/plink-binary_nwrk --recode --allow-extra-chr --noweb
plink --ped outputs/101.2/plink-binary_nwrk.ped --map outputs/101.2/plink-binary_nwrk.map --recode vcf --allow-extra-chr -out outputs/101.2/plink_nwrk
module load tabix
bgzip outputs/101.2/plink_nwrk.vcf
tabix outputs/101.2/plink_nwrk.vcf.gz
module load bcftools
bcftools reheader --samples bamlists/nwrk.names outputs/101.2/plink_nwrk.vcf.gz > outputs/101.2/renamed_nwrk.vcf.gz
module load tabix
tabix outputs/101.2/renamed_nwrk.vcf.gz
```

filtering snps in LD
```{sh, eval=FALSE}
module load bcftools
bcftools +fill-tags outputs/101.2/renamed_nwrk.vcf.gz  -- -t MAF,F_MISSING | bcftools view -i 'F_MISSING < 0.05' | bcftools view -i 'INFO/MAF > 0.05' |  bcftools +prune -m 0.20 -w 50kb --nsites-per-win 1 --nsites-per-win-mode maxAF  > outputs/101.2/pruned_nwrk.vcf

bcftools +fill-tags outputs/101.2/renamed_nwrk.vcf.gz  -- -t MAF,F_MISSING | bcftools view -i 'F_MISSING < 0.05' | bcftools view -i 'INFO/MAF > 0.05' |  bcftools +prune -m 0.20 -w 50kb --nsites-per-win 1 --nsites-per-win-mode rand  > outputs/101.2/pruned-rand_nwrk.vcf
```

moving to local storage
```{sh, eval=FALSE}
gzip -k outputs/101.2/pruned_nwrk.vcf

# switch to home terminal. not on farm
scp jsweeney@farm.hpc.ucdavis.edu:/group/ajfingergrp/RAD_projects/newark/outputs/101.2/pruned_nwrk.vcf.gz Desktop/newark/outputs/101.2
```

```{r}
snps_nwrk<-import.snpR.data(genotypes = "outputs/101.2/pruned_nwrk.vcf.gz", 
      sample.meta = m_nwrk %>% select(Pop))

pca_nwrk<-plot_clusters(snps_nwrk)
pca_nwrk$plots
```

```{r}
df_nwrk<-pca_nwrk$data$pca
ggplot(df_nwrk) +
  geom_point(aes(x=PC1, y=PC2, fill=Pop), pch=21) +
  theme_bw() +
  theme(panel.grid = element_blank()) +
  xlab(paste0("PC1 (",pca$pca_loadings[1],"%)")) +
  ylab(paste0("PC2 (",pca$pca_loadings[2],"%)")) +
  ggtitle("newarkensis only snp discovery") +
  theme(plot.title = element_text(hjust=0.5))
```


