---
title: "102.2-ML-phylogeny"
output: html_document
---

```{r}
library(ggtree)
library(ggsci)
library(tidyverse)
library(phangorn)
```
load metadata
```{r}
align496<-read_csv("outputs/100/align-496.csv", col_names = c("Ind","Reads","Dedup","Cov"))
filtered_align <- subset(align496, Dedup > 500000)
filtered_align <- subset(filtered_align, Cov > 0.1)
meta<-read_csv("metadata/meta01222025.csv")

m_fil <-left_join(filtered_align,meta, by="Ind") %>% 
  mutate(Path=paste0("data/",Ind,".sort.flt.bam")) %>% 
  mutate(Pop=`Tributary/Collection Location`) 

western_pops <- unique(m_fil$Pop[!(m_fil$Pop %in% c("NV12", "NV17", "NV18", "NV19"))])
NV17_19 <- c("NV17", "NV18", "NV19")

m_fil <- m_fil %>% mutate(Lineage = case_when(Pop %in% western_pops ~ "West", 
                                              Pop %in% NV17_19 ~ "NV17-19",
                                              Pop == "NV12" ~ "NV12"))

m_nwrk <- subset(m_fil, Species == "newarkensis" )

m_nwrk_west <- subset(m_nwrk, Pop %in% western_pops)
```

sampling max 5 inds from each site for making trimmed phylogeny later 
```{r}
set.seed(23)
m_fil_sampled <- m_fil %>%
  group_by(Pop) %>%
  slice_sample(n = 5) %>%
  ungroup()

sampled_ids <- m_fil_sampled$GVL_Code

m_nwrk_sampled <- m_nwrk %>%
  group_by(Pop) %>%
  slice_sample(n = 5) %>%
  ungroup()

sampled_nwrk_ids <- m_nwrk_sampled$GVL_Code
```

convert vcf to phylip
```{sh, eval=FALSE}
#curl -LO https://raw.githubusercontent.com/edgardomortiz/vcf2phylip/master/vcf2phylip.py
module load conda
conda activate py2
python2 vcf2phylip.py -i outputs/101.2/pruned-fil.vcf
python2 vcf2phylip.py -i outputs/101.2/pruned-nwrk.vcf
python2 vcf2phylip.py -i outputs/101.2/pruned-nwrk_west.vcf
python2 vcf2phylip.py -i outputs/101.2/pruned-rand-fil.vcf
python2 vcf2phylip.py -i outputs/101.2/pruned-rand-nwrk.vcf
python2 vcf2phylip.py -i outputs/101.2/pruned-rand-nwrk_west.vcf
#mkdir outputs/102.2
mv *.phy outputs/102.2
conda deactivate
```

run iqtree
```{sh, eval=FALSE}
#conda create -n iqtree -c bioconda -c conda-forge iqtree
module load conda
conda activate iqtree
sbatch run_iqtree.sh
```

load trees after transfering to local output 
making tree for all species in dataset
```{r}
fil_tre <- read.tree("outputs/102.2/pruned-fil.min4.phy.treefile")
t1 <- ggtree(fil_tre) + 
  geom_tiplab() + 
  geom_treescale()
t1
```

sampling for 5 inds from each site to get a cleaner tree
```{r}
# making tree with sampled inds from earlier
tips_to_drop <- fil_tre$tip.label[!fil_tre$tip.label %in% sampled_ids]
fil_tre_s <- drop.tip(fil_tre, tips_to_drop)

df_sampled <- relocate(m_fil_sampled, GVL_Code) %>% rename(label = GVL_Code)
t1s <- ggtree(fil_tre_s) %<+% df_sampled +
  geom_tippoint(aes(color = Species), cex = 3) +
  geom_tiplab(aes(label = Pop), size = 2) +
  geom_nodepoint(aes(subset = as.numeric(label) > 85), size = 1, color = "gray") +
  geom_treescale()
t1s
```

re-rooting at midpoint 
```{r}
fil_tre_s_m <- midpoint(fil_tre_s, node.labels = "support")
t1m <- ggtree(fil_tre_s_m) %<+% df_sampled +
  geom_tippoint(aes(color = Lineage), cex = 5, alpha = 0.9) +
  geom_nodepoint(aes(subset = as.numeric(label) > 95), size = 2, shape = 21, 
                 fill = "gray", color = "black") +
  geom_treescale(y = -3, fontsize = 6) + 
  scale_color_brewer(palette = "Set2")

t1m
```

making tree for just newarkensis with color by sample site 
```{r}
nwrk_tre <- read.tree("outputs/102.2/pruned-nwrk.min4.phy.treefile")
t2 <- ggtree(nwrk_tre) + 
  geom_tiplab() + 
  geom_treescale()
t2
```

sampling for 5 inds only 
```{r}
# making tree with sampled inds from earlier
tips_to_drop <- nwrk_tre$tip.label[!nwrk_tre$tip.label %in% sampled_ids]
nwrk_tre_s <- drop.tip(nwrk_tre, tips_to_drop)

df_sampled <- relocate(m_fil_sampled, GVL_Code) %>% rename(label = GVL_Code)
t2s <- ggtree(nwrk_tre_s) %<+% df_sampled +
  geom_tippoint(aes(color = Pop), cex = 3) +
  geom_tiplab(aes(label = Pop), size = 2) +
  geom_nodepoint(aes(subset = as.numeric(label) > 90), size = 1, color = "gray") +
  geom_treescale()
t2s
```

re-rooting at midpoint again
```{r}
nwrk_tre_s_m <- midpoint(nwrk_tre_s, node.labels = "support")
t2m <- ggtree(nwrk_tre_s_m) %<+% df_sampled +
  geom_tippoint(aes(color = Pop), cex = 5, alpha = 0.7) +
  geom_nodepoint(aes(subset = as.numeric(label) > 95),
                 size = 3, shape = 21, fill = "gray", color = "black") +
  geom_treescale(y = -2, fontsize = 3.5) +
  labs(color = "Sampling Site") +
  theme(legend.title    = element_text(size = 14, face = "bold"),
        legend.text     = element_text(size = 12),
        legend.key.size = unit(1.2, "lines"))
t2m
```


making tree for just newarkensis "western" cluster with color by sample site 
```{r}
nwrk_west_tre <- read.tree("outputs/102.2/pruned-nwrk_west.min4.phy.treefile")
t3 <- ggtree(nwrk_west_tre) + 
  geom_tiplab() + 
  geom_treescale()
t3
```

```{r}
# making tree with sampled inds from earlier
tips_to_drop <- nwrk_west_tre$tip.label[!nwrk_west_tre$tip.label %in% sampled_ids]
nwrk_tre_s <- drop.tip(nwrk_west_tre, tips_to_drop)

df_sampled <- relocate(m_fil_sampled, GVL_Code) %>% rename(label = GVL_Code)
t3s <- ggtree(nwrk_tre_s) %<+% df_sampled +
  geom_tippoint(aes(color = Pop), cex = 3) +
  geom_tiplab(aes(label = Pop), size = 2) +
  geom_nodepoint(aes(subset = as.numeric(label) > 90), size = 1, color = "gray") +
  geom_treescale()
t3s
```