---
title: "105.1-pop-stats-snpR"
author: "joe sweeney"
date: "2026-06-24"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

loading libs

```{r message=FALSE}
library(tidyverse)
library(snpR)
```

loading metadata

```{r message=FALSE}
align496<-read_csv("outputs/100/align-496.csv", col_names = c("Ind","Reads","Dedup","Cov"))
filtered_align <- subset(align496, Dedup > 500000)
filtered_align <- subset(filtered_align, Cov > 0.1)
meta<-read_csv("metadata/meta01222025.csv")
m_fil <-left_join(filtered_align,meta, by="Ind") %>% 
  mutate(Path=paste0("data/",Ind,".sort.flt.bam")) %>% 
  mutate(Pop=`Tributary/Collection Location`)
m_nwrk <- subset(m_fil, Species == "newarkensis" )
exclude_sites <- c("NV12", "NV17", "NV18", "NV19")
m_nwrk_west <- subset(m_nwrk, !Pop %in% exclude_sites)
```

loading snpR data and getting rid of special characters/spaces in Pop

```{r message=FALSE, warning=FALSE}
m_nwrk$Pop <- gsub(" ", "", m_nwrk$Pop)
m_nwrk$Pop <- gsub("'", "", m_nwrk$Pop)
snps<-import.snpR.data(genotypes = "outputs/101.2/pruned-nwrk.vcf", 
      sample.meta = m_nwrk %>% select(Pop))
```

calculating Tajima's D first getting chromosome size contigs from bicolor reference

```{shell eval=FALSE, include=FALSE}
cat ~/genomes/genome.fasta.fai | cut -f 1,2 | sort -k 2 -n | awk '$2 > 1e6 {print}' > chroms.txt
```

filtering snpR object for contigs from list of 'chromosomes'

```{r message=FALSE, warning=FALSE}
keep_contigs <- read.table("chroms.txt", header = FALSE)$V1
snps_chroms <- snps[which(snps@snp.meta$CHROM %in% keep_contigs), ]
```

now calculating

```{r warning=FALSE}
td_calc <- calc_tajimas_d(snps_chroms, "Pop")
td <- get.snpR.stats(td_calc, "Pop", stats = "tajimas_d")
td$weighted.means[,c(2,8)]
```

calculating nt diversity adapted from github.com/MacCampbell/lates-wgs/108-wgs-snps.Rmd

```{r}
picalc<-calc_pi(snps_chroms,"Pop")
pi<-get.snpR.stats(picalc,"Pop","pi")
pidf<-pi$single %>% as_tibble() 
pidf <- pidf %>% filter(CHROM != "JBECYN010000026.1")
head(pidf)
```

plotting pi

```{r echo=FALSE, message=FALSE}
piplot<-ggplot(pidf %>% group_by(subfacet,CHROM) %>% summarize(meanpi=mean(pi))) +
  geom_boxplot(aes(x=subfacet,y=meanpi, fill=subfacet), alpha=0.75) +
  theme_bw() +
  theme(panel.grid = element_blank()) +
  ylab("Pi\n") +
  xlab("\nSampling Site")+
  scale_x_discrete(labels=sort(unique(m_nwrk$Pop))) +
  theme(axis.text.x = element_text(angle=45, hjust=1)) +
  theme(legend.position = "") +
  scale_fill_viridis_d(option="magma") +
  theme(plot.title=element_text(size=16, face="bold"))+
  theme(axis.title = element_text(size=14, face="bold"))

piplot
```

calculating global Fst

```{r}
gfst_calc <- calc_global_fst(snps_chroms, "Pop")
gfst <- get.snpR.stats(gfst_calc, "Pop", stats = "fst")
gfst$weighted.means
```

global mean Fst \~ 0.14

calculating pairwise Fst

```{r , message=FALSE}
pwfst_calc <- calc_pairwise_fst(snps_chroms, "Pop")
pwfst <- get.snpR.stats(pwfst_calc, "Pop", stats = "fst")
```

plotting pairwise Fst on a heatmap

```{r echo=FALSE, fig.width=12, fig.height=10}
plot_pairwise_fst_heatmap(pwfst_calc, "Pop")
```
